### 2008-04-22 opal/documents/grid2008/paper.tex
added a template with the correct formatting

\documentclass[conference]{IEEEtran}

\usepackage{graphicx}

\begin{document}

% paper title
\title{On Supporting Molecular Visualization on Grid Resources using Visual
  Programming Environments}

% author information
\author{\authorblockN{Sriram Krishnan\authorrefmark{1}\authorrefmark{2},
    Guillaume Vareille\authorrefmark{3},
    Wesley Goodman\authorrefmark{1},
    Luca Clementi\authorrefmark{1},\\
    Jane Ren\authorrefmark{1},
    Sargis Dallakyan\authorrefmark{3},
    Wilfred Li\authorrefmark{1}\authorrefmark{2},
    Michel Sanner\authorrefmark{3} and
    Peter Arzberger\authorrefmark{1}}
  \authorblockA{\authorrefmark{1}National Biomedical Computation Resource\\
    CRBS, UC San Diego, 9500 Gilman Dr, La Jolla, CA 92093}
  \authorblockA{\authorrefmark{2}San Diego Supercomputer Center\\
    UC San Diego MC 0505, 9500 Gilman Dr, La Jolla, CA 92093}
  \authorblockA{\authorrefmark{3}The Molecular Graphics Lab\\
    The Scripps Research Institute, La Jolla, CA 92093
    }
  \authorblockA{sriram@sdsc.edu, vareille@scripps.edu, \{wgoodman,
      lclement, j2ren@ucsd.edu\},\\
      wilfred@sdsc.edu, sanner@scripps.edu, parzberger@ucsd.edu}
  }

% make the title area
\maketitle

\begin{abstract}

  Abstract goes here (Sriram)

\end{abstract}

\section{Intro on scientific workflows and grid computing}

- Scientific workflows: Michel

- Grid computing: Sriram

\section {Vision vs other workflow management systems (MS, except section
  2.c)}

- Design -- workflow representation and control: Michel

- API -- how to write Vision nodes, introspection, etc: Michel

- Comparison with other tools: Luca

  - Kepler, Ptolemy and MoML

  - Taverna, Scufl and Freeflou

\section {Application based services}

- Opal wrapper and GUI \cite{opal}: Sriram

- Opal and Opal-CSF4: Sriram

- Data services: Sriram

\section {Service based workflows}

- Vision based grid workflows (Michel, Guillaume, Sargis)

- Virtualization of Opal services (Sargis - something about how you
automatically create Vision nodes from the service list)

- Virtualization of Vision based workflows

  i) Use a local CLI opal client to access local or remote services (Sriram)

 ii) Automatic vision node creation through the opal client information
 service (Sriram, Guillaume - how the generated class is converted into a Vision node using introspection)

\section {Discussion - Workflow modeling and execution}

- How vision currently stores workflow (Michel)

- Fault tolerance (Michel, Sriram)

- Support for XPDL and other representations for workflows (Luca)

\section {Conclusions}

- Conclusions (Sriram)

%% References
\bibliographystyle{plain}
\bibliography{paper}

% that's all folks
\end{document}

### 2008-04-22 src/edu/sdsc/nbcr/opal/AppServiceImpl.java
fixed a bug where a NPE was being thrown while waiting for job to activate

@@ -514,7 +514,7 @@ public class AppServiceImpl
 	// wait for job activation
 	StatusOutputType status = null;
 	try {
-	    jobManager.waitForActivation();
+	    status = jobManager.waitForActivation();
 	} catch (JobManagerException jme) {
 	    logger.error(jme.getMessage());
 

### 2008-04-22 opal/opal-core/src/edu/sdsc/nbcr/opal/AppServiceImpl.java
added code to read in another property for absolute path of working dir

@@ -149,6 +149,13 @@ public class AppServiceImpl
 	try {
 	    File prefixDir = new File(outputPrefix);
 	    outputPrefix = prefixDir.getCanonicalPath();
+	    String absolutePath = props.getProperty("absolute.path");
+	    if (absolutePath != null) {
+		outputPrefix = absolutePath;
+		logger.info("Found absolute path for working dir: " + outputPrefix);
+	    } else {
+		logger.info("Found canonical path for working dir: " + outputPrefix);
+	    }
 	} catch (IOException e) {
 	    logger.fatal(e);
 	}

### 2008-04-23 opal/documents/grid2008/paper.tex
added abstract

@@
 
-  Abstract goes here (Sriram)
+  The visual programming paradigm provided by tools such as AVS [ref],
+  OpenDX [ref], SciRun [ref], etc have been around for a couple decades.
+  Recently, with the advent of the Web 2.0 paradigm and the concept of {\em
+    mashups} [ref], there has been renewed interest in such types of
+  environments. Over the past several years, we have developed one such
+  tool called Vision [ref], which is a component-based, application-domain
+  agnostic, cross-platform visual programming environment. Meanwhile,
+  advances in distributed and Grid computing have opened up avenues for
+  scientists to run large simulations on a scale that was impossible until
+  recent times. In the past, visual programming tools were generally
+  limited to running scientific experiments on user's individual
+  workstations, and had limited capability to use distributed resources.
+  With the proliferation of Web services, distributed resources are much
+  more easy to access from any workstation that is connected to the
+  Internet. In particular, we have developed the Opal toolkit, which
+  provides a mechanism to wrap scientific applications running on Grid
+  resources as Web services, thus providing the clients with a simple Web
+  API to the back-end computational functionality. Opal encapsulates
+  standard Grid security mechanisms, schedulers, and state management for
+  jobs, and also provides a mechanism to describe command-line arguments
+  that can be used to automatically generate user-interfaces [ref]. In this
+  paper, we describe our current work on enabling access to scientific
+  applications running on Grid resources via Opal from the Vision
+  environment. We present brief overviews of both Vision and Opal, and
+  discuss a scientific workflow that has been implemented with the above
+  tools. We discuss our experiences, and plans for future work.
 

### 2008-04-23 opal/documents/grid2008/paper.tex
added Michel and Luca's initial text

@@
 
+The visual-programming paradigm provided by tools such as the AVS [X],
+OpenDX [X], Khoros [X], DataExplorer [X], Ptoleme [X] or Scirun, [X] has
+been around for a couple of decades. Recently, there has recently been a
+renewed interest in this type of user interface and numerous pipelining
+tools have become available, including: PHENIX [x], pipeline pilot [X],
+mapero [X], Orange [X], Elefant [X], Kepler [X], Taverna [X] to name a few.
+
+These software environments, allow a user to construct a computational
+network by placing nodes performing a specific task onto a canvas and
+connecting their output and input ports using pipes that specify the flow
+of data through the network. The appeal of this type of interface is that
+the user can interactively construct networks of computational nodes that
+correspond to simple programs without having to write a single line of
+code. It provides an effective way to hide syntax details and data
+structures used to store data from the user who in general cares more about
+the result of the computation rather than these implementation details. In
+other words it empowers users with an unprecedented level of flexibility in
+using computers without having to become a programmer.
+
+While some of these tools are highly specific to a given application
+domain, other try to be more generic programming environments. However,
+most of these programs are what we call “application-centric” (i.e. they
+constitute the development environment for a specific application) rather
+than being a software component that will integrate and dovetail nicely
+with other software components and that does not make assumptions about the
+domain of application.
+
 - Grid computing: Sriram
 
-\section {Vision vs other workflow management systems (MS, except section
-  2.c)}
+\section {Background}
 
-- Design -- workflow representation and control: Michel
+\subsection {The Vision Environment}
 
-- API -- how to write Vision nodes, introspection, etc: Michel
+% - Design -- workflow representation and control: Michel
 
-- Comparison with other tools: Luca
+% - API -- how to write Vision nodes, introspection, etc: Michel
 
-  - Kepler, Ptolemy and MoML
+Over the past 8 years we have developed Vision [X], a component-based,
+application-domain agnostic, cross-platform visual programming environment.
+Vision has been designed as an extension to the Python programming language
+[X] thus providing a fully fledged, high level, object-oriented,
+interpretive language for the interactive manipulation of data and/or
+modification of the application itself.
 
-  - Taverna, Scufl and Freeflou
+The Vision framework was designed to dynamically load libraries of
+computational nodes providing specific functionality. Not only does this
+approach reduce the footprint of Vision in memory, but it also avoids
+cluttering the interface with nodes that a particular user might not be
+interested in using. Each library can define its own data types, allowing
+the validation of data passed to a node before the node will execute. A
+data type definition includes a description of the visual representation of
+ports receiving or sending this data type, thus providing visual clues to
+the user while he create the data flow by connecting output ports to input
+ports. Port can be declared as required or optional. A node can only run if
+all required ports present valid data. Introspection tools allow the user
+to inspect data on input ports and output ports in real time.
+
+Vision network nodes are Python objects that provide input ports and output
+ports. When a node is scheduled for execution, it will collect the data on
+its input ports and pass it ot the node’s compute function. This function
+typically imports Python code from some other Python package to operate on
+the data and produce a result, which is then sent to the node’s output
+port. This important design features makes Vision “just another” user
+interface to functionality available to any program running a Python
+interpreter thus promoting code re-use and inter-operability. Moreover,
+since the node’s compute function is written in the Python programming
+language it can be inspected and modified interactively by the user.  In
+fact, a node can be designed interactively in Vision using the node editor.
+All modifications made to a node or a network, are saved in the network
+description file. This file contains Python code that will recreate the
+network when executed. We made this choice in order to avoid having to
+define a format for describing networks and also to have access to the full
+power of a programming language for the definition of these networks.
+
+Vision supports the concept of MacroNodes which appear as a simple node in
+a network, but encapsulate a sub-network. MacroNodes can exist inside
+Macronodes, thus allowing a hierarchical visual representation of algorithm
+and facilitating code reuse within Vision.
+
+While execution of a node is typically triggered by new data presented on
+its ports, a node can also be triggered by events such as the completion of
+another node’s execution. Vision’s execution model allows for the partial
+execution of networks. When a parameter changes or new data is provided for
+a given node, only the sub-graph of children nodes depending on this
+parameter or dataset will re-compute.
+
+An iterate node supports the implementation of “for loops” in Vision. This
+node loops over any sequence of Python objects and triggers the sub-network
+of its children nodes for each one of the objects in the sequence. Begin
+and end of loop events allow the accumulation of data during a loop and the
+execution of code after the completion of the loop.
+
+To date, several libraries of nodes are available [Fig X]. The standard
+library comprises nodes exposing Python keywords (print, eval, setattr and
+call methods), along with nodes enabling the generation of simple input
+data such as floats, integers, and strings. The 3-D Visualization library
+is based on the DejaVu component and provides nodes to create 3-D
+geometries and display them. The SymServ library implements a set of nodes
+defining geometric transformations such as symmetries, translation,
+rotation, helical arrangements etc. that can be combined to created streams
+of nested 3-D transformations. The MolKit library is based on the MolKit
+package and provides nodes to read molecular data files and manipulate
+molecules. The Imaging library currently exposes a subset of the
+functionality available in the Python Image Library (PIL) thus enabling the
+creation of image processing pipelines.
+
+So far, Vision has essentially been used in an interactive way with most
+computational nodes running locally (i.e. in the same process as the Python
+interpreter running Vision). In this paper we describe the addition of
+support for nodes running web services.
+
+% - Comparison with other tools: Luca
+
+%   - Kepler, Ptolemy and MoML
+
+%   - Taverna, Scufl and Freeflou
+
+\subsection {Related Work}
+
+Several tools have been developed by the scientific community to support
+the design and execution of workflows. Each one of them use different
+approaches to model a graphical environment for designing and executing
+workflow.  In this section we present three workflow management systems:
+Kepler, Taverna, and AVS. This paragraph does not want to provide an
+exaustive survey of these systems, for more information refer to [grid].
+
+Kepler [6] is a workflow system developed under several research projects.
+It is written in Java and based on Ptolemy II [6], a system originally
+created to study modeling, simulation, and design of concurrent systems.
+Kepler models workflow in MoMl, a markup language developed in Ptolemy that
+is based on the concept of actors. Each actor has input and output ports
+and wraps typical workflow actions like invoke a Web Service or a Grid
+Services, launch a Globus Grid Jobs, perform a GridFTP transfer, and many
+others. The designer can model non Directed Acyclic Graph [5] using a
+simple Graphic User Interface where the user has simply to drag and drop
+icons and connect them.  Kepler can also be classified as a workflow system
+based on the data flow model [6], where data is streamed from one actor to
+another triggering their activation. Moreover an extensible architecture
+based on the plugin model allows developers to easily interagte thrid party
+tools.
+
+Taverna is an open source software tool for designing and executing
+bioinformatics workflows, developed in the myGrid project [7]. It uses
+Scufl as a runtime language for the execution of the workflow, and Freefluo
+as an execution engine. Scufl can be extended using Processor plugins that
+manage the interaction with different external service interfaces. Taverna
+is shipped with a set of processors that guarantee interaction with
+services based on WS-I standard. The designer is based on a directed ciclic
+graph that represents data flow, which uses a simple graphical notation
+developed by the Taverna team. The GUI can be easily extended creating new
+palettes and new widgets with development of new plugins.
 

### 2008-04-24 opal/documents/grid2008/paper.tex
added conclusions. now only missing the rest of the paper :)

@@
 
+\subsection {Opal Overview}
+
 - Opal wrapper and GUI \cite{opal}: Sriram 
+- Opal and Opal-CSF4: Sriram
+- Data services: Sriram
 
-- Opal and Opal-CSF4: Sriram
-
-- Data services: Sriram
+\subsection {Towards Opal2}
 
@@
 
+- PDB2PQR-APBS workflow in action
+
 \section {Discussion - Workflow modeling and execution}
@@
 
-- Conclusions (Sriram)
+Visual programming environments enable non-programmers to intuitively and
+interactively build scientific workflows and visualize their data, without
+having to write code or understand the details of the back-end
+infrastructure. In this paper, we described our tool called Vision, which
+is a component-based, application-domain agnostic, cross-platform visual
+programming environment. On the other hand, Grid computing provides an
+enormous opportunity to scientific end-users to use distributed computing
+and storage resources to enable novel science hitherto impossible using
+traditional computing mechanisms. However, most Grid middleware tools are
+complicated to install and too jargon rich for scientists to learn. In
+order to simplify the use of Grid infrastructures, one has to provide easy
+to use abstractions to the complex back-end functionality. Towards this
+end, we have developed the Opal toolkit, which enables wrapping scientific
+applications running on Grid resources as simple Web services, which can be
+accessed via a number of user interfaces. In this paper, we described how
+we can use the metadata provided by the Opal services to automatically
+generate Vision nodes to access the applications running on Grid resources.
+With the help of a molecular visualization workflow, we demonstrated how
+this is being used by domain scientists, while being completely shielded
+from the complexity of the back-end computational infrastructure. We
+believe that the use of services-oriented science within visual programming
+environments will go a long way towards lowering the entrance barrier for
+scientific end-users to effectively use Grid middleware.
+
+The authors would like to acknowledge support from the NIH P41 RR 08605 to
+NBCR; TATRC W81XWH-07-2-0014 to PWA, the Gordon and Betty Moore Foundation
+grant for the CAMERA project, and the NSF grants INT-0314015 and
+OCI-0627026 for the PRAGMA project.
 

### 2008-04-24 opal/documents/grid2008/paper.tex
added a couple of diagrams and some text - more coming

@@
 
-\section {Background}
+\section {Visual Programming Environments}
 
@@
 
-\section {Application based services}
+\section {Services Oriented Science}
 
-\subsection {Opal Overview}
+Programming and using distributed Grid resources is quite a challenge task.
+There is hope that Service-Oriented Architectures (SOA) can address this
+problem by providing platform-independent, language-neutral service
+interfaces that hide the complexity of the implementations while providing
+a well-defined and high-performance Quality of Service (QoS). In our recent
+paper [ref], we claimed that in order to lower the entrance barrier for
+scientific application developers and end-users to Grid infrastructures, we
+must do the following:
 
-- Opal wrapper and GUI \cite{opal}: Sriram 
+% TODO: possibly expand every bullet
+\begin{itemize}
+\item{Focus on application-level services}
+\item{Provide users access to virtualized resources}
+\item{Provide multiple user interfaces to scientific users}
+\item{Lower deployment costs}
+\item{Follow the Web 2.0 paradigm by enabling user participation and contribution}
+\end{itemize}
+
+Towards this end, we have developed the Opal toolkit
+\cite{opal} for wrapping scientific applications as Web services. In order
+to wrap an application as a Web service, application developers are
+expected to write a simple XML-based application configuration. The
+application configuration contains information about the scientific
+application, such as application name, binary location, and application
+metadata (e.g. usage information). Using a simple Apache Ant task, the Opal
+toolkit can deploy the application as a Web service into a container based
+on Apache Tomcat and Axis. Once the application is deployed as a Web
+service, clients can access this service programmatically using its WSDL
+description. The WSDL API provides operations for job launch (which accepts
+command-line arguments and input files as its parameters), querying status,
+and retrieving outputs. Furthermore, it also provides an API for retrieving
+application metadata. WSDL savvy users could write their own clients to
+access Opal based applications. However, not every scientific user is
+capable or interested in writing Web service clients. Hence, several
+interfaces have been provided for the end-user. Apart from command-line
+clients written in languages like Java, Python [4] and Perl [15] available
+from NBCR, Web interfaces for Opal based services are available through the
+GridSphere portal environment [18], such as the MEME portlet in My
+WorkSphere [24]. In addition, Opal based interfaces via Rich Internet
+Applications (RIA) such as Gemstone [20], and Problem Solving Environments
+(PSE) such as the Python Molecular Viewer (PMV) [22], AutoDockTools [21]
+and Continuity [27] have also been provided.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=8cm]{architecture}
+\end{center}
+\caption{Services Oriented Science Using the Opal Toolkit}
+\label{fig:arch}
+\end{figure}
+
+Although a detailed description of the Opal toolkit is beyond the scope of
+this paper (and can be found in \cite{opal}), the overall architecture of
+is shown in Figure \ref{fig:arch}. The salient features of Opal are as
+follows:
+
+\begin{itemize}
+\item{Scheduling and Cluster Management}
+\item{Data management and Persistence}
+\item{Security}
+\item{Standardized WSDL API}
+\item{Argument Description and Dynamic Interface Generation}
+\end{itemize}
+
+- Opal wrapper and GUI: Sriram
+
 - Opal and Opal-CSF4: Sriram
+
 - Data services: Sriram
 
-\subsection {Towards Opal2}
+- Status, more information
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=8cm]{opal-gui}
+\end{center}
+\caption{Service Invocation via the Opal GUI}
+\label{fig:opal-gui}
+\end{figure}
+
+- Towards Opal2
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=8.5cm]{opal2-providers}
+\end{center}
+\caption{The Opal2 Provider Model}
+\label{fig:opal2-providers}
+\end{figure}
 

### 2008-04-24 opal/documents/grid2008/paper.tex
added a few more diagrams

@@
 
-\section{Intro on scientific workflows and grid computing}
+\section{Introduction}
 
 - Scientific workflows: Michel
+
+% TODO: Add more context
 
@@
 domain of application.
+
+% TODO: Add Grid stuff
 
@@
 
+% TODO: Need a more useful screenshot
+\begin{figure}
+\begin{center}
+\includegraphics[width=8cm]{vision-overview}
+\end{center}
+\caption{A typical scientific workflow composed using the Vision interface}
+\label{fig:vision-overview}
+\end{figure}
+
+\subsection {Related Work}
+
 % - Comparison with other tools: Luca
@@
 %   - Taverna, Scufl and Freeflou
-
-\subsection {Related Work}
 
@@
 
-% TODO: possibly expand every bullet
-\begin{itemize}
+% TODO: possibly expand every bullet (04/23)
+\begin{enumerate}
 \item{Focus on application-level services}
@@
 \item{Follow the Web 2.0 paradigm by enabling user participation and contribution}
-\end{itemize}
+\end{enumerate}
 
@@
 \begin{center}
-\includegraphics[width=8cm]{architecture}
+\includegraphics[width=8.5cm]{architecture}
 \end{center}
-\caption{Services Oriented Science Using the Opal Toolkit}
+\caption{Services Oriented Science using the Opal Toolkit}
 \label{fig:arch}
@@
 
-\begin{itemize}
+% TODO: Expand (04/23)
+\begin{enumerate}
 \item{Scheduling and Cluster Management}
@@
 \item{Argument Description and Dynamic Interface Generation}
-\end{itemize}
+\end{enumerate}
 
@@
 \begin{center}
-\includegraphics[width=8cm]{opal-gui}
+\includegraphics[width=8.5cm]{opal-gui}
 \end{center}
@@
 
-\section {Service based workflows}
+\section {Services-based Workflows in Vision}
 
@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[width=8cm]{vision-gui}
+\end{center}
+\caption{Automatic interface generation for the PDB2PQR service in Vision}
+\label{fig:vision-gui}
+\end{figure}
+
 - PDB2PQR-APBS workflow in action
 
-\section {Discussion - Workflow modeling and execution}
+% TODO: Need a more accurate screenshot
+\begin{figure}
+\begin{center}
+\includegraphics[width=8cm]{vision-workflow}
+\end{center}
+\caption{Vision worklow using local and remote services}
+\label{fig:vision-workflow}
+\end{figure}
+
+\section {Discussion}
 
@@
 %% References
+% TODO (04/23)
 \bibliographystyle{plain}

### 2008-04-24 opal/documents/grid2008/paper.tex
added a bunch of references

@@
 
-  The visual programming paradigm provided by tools such as AVS [ref],
-  OpenDX [ref], SciRun [ref], etc have been around for a couple decades.
-  Recently, with the advent of the Web 2.0 paradigm and the concept of {\em
-    mashups} [ref], there has been renewed interest in such types of
-  environments. Over the past several years, we have developed one such
-  tool called Vision [ref], which is a component-based, application-domain
-  agnostic, cross-platform visual programming environment. Meanwhile,
-  advances in distributed and Grid computing have opened up avenues for
-  scientists to run large simulations on a scale that was impossible until
-  recent times. In the past, visual programming tools were generally
-  limited to running scientific experiments on user's individual
-  workstations, and had limited capability to use distributed resources.
-  With the proliferation of Web services, distributed resources are much
-  more easy to access from any workstation that is connected to the
-  Internet. In particular, we have developed the Opal toolkit, which
-  provides a mechanism to wrap scientific applications running on Grid
-  resources as Web services, thus providing the clients with a simple Web
-  API to the back-end computational functionality. Opal encapsulates
-  standard Grid security mechanisms, schedulers, and state management for
-  jobs, and also provides a mechanism to describe command-line arguments
-  that can be used to automatically generate user-interfaces [ref]. In this
-  paper, we describe our current work on enabling access to scientific
-  applications running on Grid resources via Opal from the Vision
-  environment. We present brief overviews of both Vision and Opal, and
-  discuss a scientific workflow that has been implemented with the above
-  tools. We discuss our experiences, and plans for future work.
+  The visual programming paradigm provided by tools such as AVS \cite{avs},
+  OpenDX \cite{opendx}, SciRun \cite{scirun}, etc have been around for a
+  couple decades. Recently, with the advent of the Web 2.0 paradigm and the
+  concept of {\em mashups} \cite{web2.0}, there has been renewed interest
+  in such types of environments. Over the past several years, we have
+  developed one such tool called Vision \cite{viper}, which is a
+  component-based, application-domain agnostic, cross-platform visual
+  programming environment. Meanwhile, advances in distributed and Grid
+  computing have opened up avenues for scientists to run large simulations
+  on a scale that was impossible until recent times. In the past, visual
+  programming tools were generally limited to running scientific
+  experiments on user's individual workstations, and had limited capability
+  to use distributed resources. With the proliferation of Web services,
+  distributed resources are much more easy to access from any workstation
+  that is connected to the Internet. In particular, we have developed the
+  Opal toolkit, which provides a mechanism to wrap scientific applications
+  running on Grid resources as Web services, thus providing the clients
+  with a simple Web API to the back-end computational functionality. Opal
+  encapsulates standard Grid security mechanisms, schedulers, and state
+  management for jobs, and also provides a mechanism to describe
+  command-line arguments that can be used to automatically generate
+  user-interfaces \cite{opal}. In this paper, we describe our current work
+  on enabling access to scientific applications running on Grid resources
+  via Opal from the Vision environment. We present brief overviews of both
+  Vision and Opal, and discuss a scientific workflow that has been
+  implemented with the above tools. We discuss our experiences, and plans
+  for future work.
 
@@
 
-The visual-programming paradigm provided by tools such as the AVS [X],
-OpenDX [X], Khoros [X], DataExplorer [X], Ptoleme [X] or Scirun, [X] has
-been around for a couple of decades. Recently, there has recently been a
-renewed interest in this type of user interface and numerous pipelining
-tools have become available, including: PHENIX [x], pipeline pilot [X],
-mapero [X], Orange [X], Elefant [X], Kepler [X], Taverna [X] to name a few.
+The visual-programming paradigm provided by tools such as the AVS
+\cite{avs}, OpenDX \cite{opendx}, or SciRun \cite{scirun} has been around
+for a couple of decades. Recently, there has recently been a renewed
+interest in this type of user interface and numerous pipelining tools have
+become available, including: PHENIX [ref], pipeline pilot [ref], mapero
+[ref], Orange [ref], Elefant [ref], Kepler \cite{kepler}, Taverna
+\cite{taverna} to name a few.
 
@@
 
-Over the past 8 years we have developed Vision [X], a component-based,
-application-domain agnostic, cross-platform visual programming environment.
-Vision has been designed as an extension to the Python programming language
-[X] thus providing a fully fledged, high level, object-oriented,
-interpretive language for the interactive manipulation of data and/or
-modification of the application itself.
+Over the past 8 years we have developed Vision \cite{viper}, a
+component-based, application-domain agnostic, cross-platform visual
+programming environment. Vision has been designed as an extension to the
+Python programming language thus providing a fully fledged, high level,
+object-oriented, interpretive language for the interactive manipulation of
+data and/or modification of the application itself.
 
@@
 
-To date, several libraries of nodes are available [Fig X]. The standard
-library comprises nodes exposing Python keywords (print, eval, setattr and
-call methods), along with nodes enabling the generation of simple input
-data such as floats, integers, and strings. The 3-D Visualization library
-is based on the DejaVu component and provides nodes to create 3-D
-geometries and display them. The SymServ library implements a set of nodes
-defining geometric transformations such as symmetries, translation,
-rotation, helical arrangements etc. that can be combined to created streams
-of nested 3-D transformations. The MolKit library is based on the MolKit
-package and provides nodes to read molecular data files and manipulate
-molecules. The Imaging library currently exposes a subset of the
-functionality available in the Python Image Library (PIL) thus enabling the
-creation of image processing pipelines.
+To date, several libraries of nodes are available as seen in Figure
+\ref{fig:vision-overview}. The standard library comprises nodes exposing
+Python keywords (print, eval, setattr and call methods), along with nodes
+enabling the generation of simple input data such as floats, integers, and
+strings. The 3-D Visualization library is based on the DejaVu component and
+provides nodes to create 3-D geometries and display them. The SymServ
+library implements a set of nodes defining geometric transformations such
+as symmetries, translation, rotation, helical arrangements etc. that can be
+combined to created streams of nested 3-D transformations. The MolKit
+library is based on the MolKit package and provides nodes to read molecular
+data files and manipulate molecules. The Imaging library currently exposes
+a subset of the functionality available in the Python Image Library (PIL)
+thus enabling the creation of image processing pipelines.
 
@@
 Kepler, Taverna, and AVS. This paragraph does not want to provide an
-exaustive survey of these systems, for more information refer to [grid].
+exaustive survey of these systems, for more information refer to
+\cite{wf-grid}.
 
-Kepler [6] is a workflow system developed under several research projects.
-It is written in Java and based on Ptolemy II [6], a system originally
-created to study modeling, simulation, and design of concurrent systems.
-Kepler models workflow in MoMl, a markup language developed in Ptolemy that
-is based on the concept of actors. Each actor has input and output ports
-and wraps typical workflow actions like invoke a Web Service or a Grid
-Services, launch a Globus Grid Jobs, perform a GridFTP transfer, and many
-others. The designer can model non Directed Acyclic Graph [5] using a
+Kepler is a workflow system developed under several research projects. It
+is written in Java and based on Ptolemy II \cite{ptolemy}, a system
+originally created to study modeling, simulation, and design of concurrent
+systems. Kepler models workflow in MoMl, a markup language developed in
+Ptolemy that is based on the concept of actors. Each actor has input and
+output ports and wraps typical workflow actions like invoke a Web Service
+or a Grid Services, launch a Globus Grid Jobs, perform a GridFTP transfer,
+and many others. The designer can model non Directed Acyclic Graph using a
 simple Graphic User Interface where the user has simply to drag and drop
 icons and connect them. Kepler can also be classified as a workflow system
-based on the data flow model [6], where data is streamed from one actor to
-another triggering their activation. Moreover an extensible architecture
-based on the plugin model allows developers to easily interagte thrid party
-tools.
+based on the data flow model \cite{kepler}, where data is streamed from one
+actor to another triggering their activation. Moreover an extensible
+architecture based on the plugin model allows developers to easily
+interagte thrid party tools.
 
 Taverna is an open source software tool for designing and executing
-bioinformatics workflows, developed in the myGrid project [7]. It uses
-Scufl as a runtime language for the execution of the workflow, and Freefluo
-as an execution engine. Scufl can be extended using Processor plugins that
-manage the interaction with different external service interfaces. Taverna
-is shipped with a set of processors that guarantee interaction with
-services based on WS-I standard. The designer is based on a directed ciclic
-graph that represents data flow, which uses a simple graphical notation
-developed by the Taverna team. The GUI can be easily extended creating new
-palettes and new widgets with development of new plugins.
+bioinformatics workflows, developed in the myGrid project \cite{taverna}.
+It uses Scufl as a runtime language for the execution of the workflow, and
+Freefluo as an execution engine. Scufl can be extended using Processor
+plugins that manage the interaction with different external service
+interfaces. Taverna is shipped with a set of processors that guarantee
+interaction with services based on WS-I standard. The designer is based on
+a directed ciclic graph that represents data flow, which uses a simple
+graphical notation developed by the Taverna team. The GUI can be easily
+extended creating new palettes and new widgets with development of new
+plugins.
 
@@
 a well-defined and high-performance Quality of Service (QoS). In our recent
-paper [ref], we claimed that in order to lower the entrance barrier for
-scientific application developers and end-users to Grid infrastructures, we
-must do the following:
+paper \cite{escience}, we claimed that in order to lower the entrance
+barrier for scientific application developers and end-users to Grid
+infrastructures, we must do the following:
 
-% TODO: possibly expand every bullet (04/23)
 \begin{enumerate}
-\item{Focus on application-level services}
-\item{Provide users access to virtualized resources}
-\item{Provide multiple user interfaces to scientific users}
-\item{Lower deployment costs}
-\item{Follow the Web 2.0 paradigm by enabling user participation and contribution}
+\item{{\bf Focus on application-level services:} From the perspective of a
+    domain scientist, the services that are most relevant are the ones that
+    provide a scientific operation, and where the semantics of the
+    operations are defined in terms of the domain science, and not computer
+    science. This is what Foster terms ''services-oriented science''
+    \cite{sos}.}
+
+\item{{\bf Provide users access to virtualized resources:} The use of
+    virtualized resources through a layer of indirection allows the system
+    developers to add or move resources as needed without any interruption.}
+
+\item{{\bf Provide multiple user interfaces to scientific users:} A single
+    user interface is typically not sufficient for all the users using Grid
+    resources. We advocate application services that can be used by a
+    number of end-user tools, including Web-based portals, rich desktop
+    applications, command-line tools and language-specific libraries.}
+
+\item{{\bf Lower deployment costs:} Scientific end-users don’t have the
+    time or money to invest in trying to figure out the complexity of the
+    back-end resources, and deploying an application on Grid resources
+    should be as easy as possible.}
+
+\item{{\bf Follow the Web 2.0 paradigm by enabling user participation and
+      contribution:} In principle, the Web 2.0 paradigm allows users to
+    contribute their applications, share data, and participate in
+    collaborative activities with ease and keep their focus on the process
+    of creation, instead of maintaining and learning new tools and
+    interfaces.}
 \end{enumerate}
 
-Towards this end, we have developed the Opal toolkit
-\cite{opal} for wrapping scientific applications as Web services. In order
-to wrap an application as a Web service, application developers are
-expected to write a simple XML-based application configuration. The
-application configuration contains information about the scientific
-application, such as application name, binary location, and application
-metadata (e.g. usage information). Using a simple Apache Ant task, the Opal
-toolkit can deploy the application as a Web service into a container based
-on Apache Tomcat and Axis. Once the application is deployed as a Web
-service, clients can access this service programmatically using its WSDL
-description. The WSDL API provides operations for job launch (which accepts
-command-line arguments and input files as its parameters), querying status,
-and retrieving outputs. Furthermore, it also provides an API for retrieving
-application metadata. WSDL savvy users could write their own clients to
-access Opal based applications. However, not every scientific user is
-capable or interested in writing Web service clients. Hence, several
-interfaces have been provided for the end-user. Apart from command-line
-clients written in languages like Java, Python [4] and Perl [15] available
-from NBCR, Web interfaces for Opal based services are available through the
-GridSphere portal environment [18], such as the MEME portlet in My
-WorkSphere [24]. In addition, Opal based interfaces via Rich Internet
-Applications (RIA) such as Gemstone [20], and Problem Solving Environments
-(PSE) such as the Python Molecular Viewer (PMV) [22], AutoDockTools [21]
-and Continuity [27] have also been provided.
+Towards this end, we have developed the Opal toolkit \cite{opal} for
+wrapping scientific applications as Web services. In order to wrap an
+application as a Web service, application developers are expected to write
+a simple XML-based application configuration. The application configuration
+contains information about the scientific application, such as application
+name, binary location, and application metadata (e.g. usage information).
+Using a simple Apache Ant task, the Opal toolkit can deploy the application
+as a Web service into a container based on Apache Tomcat and Axis. Once the
+application is deployed as a Web service, clients can access this service
+programmatically using its WSDL description. The WSDL API provides
+operations for job launch (which accepts command-line arguments and input
+files as its parameters), querying status, and retrieving outputs.
+Furthermore, it also provides an API for retrieving application metadata.
+WSDL savvy users could write their own clients to access Opal based
+applications. However, not every scientific user is capable or interested
+in writing Web service clients. Hence, several interfaces have been
+provided for the end-user. Apart from command-line clients written in
+languages like Java, Python and Perl available from NBCR, Web interfaces
+for Opal based services are available through the GridSphere portal
+environment \cite{gridsphere}, such as the MEME portlet in My WorkSphere
+\cite{myworksphere}. In addition, Opal based interfaces via Rich Internet
+Applications (RIA) such as Gemstone \cite{gemstone}, and Problem Solving
+Environments (PSE) such as the Python Molecular Viewer (PMV), AutoDockTools
+\cite{pmv-vision-adt} and Continuity \cite{continuity} have also been provided.
 
@@
 \begin{center}
-\includegraphics[width=8.5cm]{opal2-providers}
+\includegraphics[width=9cm]{opal2-providers}
 \end{center}
@@
 %% References
-% TODO (04/23)
 \bibliographystyle{plain}

### 2008-04-24 opal/documents/grid2008/paper.tex
added more details to Opal section

@@
 
-% TODO: Expand (04/23)
 \begin{enumerate}
-\item{Scheduling and Cluster Management}
-\item{Data management and Persistence}
-\item{Security}
-\item{Standardized WSDL API}
-\item{Argument Description and Dynamic Interface Generation}
+\item{{\bf Scheduling and Cluster Management:} Since different sites use
+    different schedulers such as SGE \cite{sge}, Condor \cite{condor},
+    etc., Opal provides a uniform way to access them via standard APIs such
+    as Globus GRAM \cite{gram} and DRMAA \cite{drmaa}. The schedulers are
+    simply specified using a properties file, and Opal configures itself
+    appropriately to access the schedulers via the appropriate parameters.
+    In addition, we are currently also able to submit jobs to the CSF4
+    metascheduler \cite{csf4} using its WSRF API \cite{wsrf} as described
+    in \cite{opal-gui}.}
+
+\item{{\bf Data management and Persistence:} The Opal toolkit manages the
+    job data for a particular run on behalf of the user. Every job is run
+    in a separate workspace so that multiple jobs from different users can
+    be run concurrently. Furthermore, information about job status and
+    outputs is persisted in a database to provide fault-tolerance for the
+    services.}
+
+\item{{\bf Security:} Opal services can be configured to use
+    transport-level GSI-based \cite{gsi} security so as to restrict access
+    to only authorized users.}
+
+\item{{\bf Standardized WSDL API:} All Opal use a standard WSDL so that
+    they can accessed in a uniform manner by different clients. We choose
+    not to generate application specific WSDLs to avoid regneration of
+    client-side stubs and rewriting of clients for every application.}
+
+\item{{\bf Argument Description and Dynamic Interface Generation:} Opal
+    services can be optionally configured with an XML specification for the
+    command-line arguments. This specification consists of a description
+    for flags, tagged and untagged parameters, and grouping of arguments.
+    Flags are not ordered and are usually represented with a character
+    prefixed with a dash and they activate a functionality in the
+    application (e.g. -verbose). Tagged Parameters are usually formed by a
+    prefix and input value (e.g. -input <filename>) - they can appear in
+    any order. Untagged parameters are not prefixed, and hence their order
+    is relevant. The argument description is mainly useful for two purposes
+    - for validation of command-line arguments passed to the application,
+    and for automatic interface generation. A detailed description of the
+    argument description, and automatic generation of Web forms from the
+    description is provided in \cite{opal-gui}. A complete lifecycle of
+    this process is shown in Figure \ref{fig:opal-gui}.}
+
 \end{enumerate}
-
-- Opal wrapper and GUI: Sriram
-
-- Opal and Opal-CSF4: Sriram
-
-- Data services: Sriram
-
-- Status, more information
 
@@
 \end{figure}
+
+The Opal toolkit is freely available from SourceForge, and uses an Open
+Source BSD-style license. The current stable version of Opal at the time of
+writing this manuscript is 1.9.3, and more information is available from
+out website \cite{opal-website}.
 
